Hybrid Threads for the Parrot Virtual Machine

### 2012-06-23 Paper.tex
Second round of fixes suggested by H. Dobler++

This thesis shows a way to implement threading support in Parrot using a hybrid approach using a combination of lightweight (``green'') threads and operating system threads. These lightweight threads are used as messages in a system where reading shared data is allowed but only the one owner thread may write to it. The implementation of this hybrid approach is described in detail and benchmarks are presented demonstrating the viability of this design.

@@

In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

@@

A mark and sweep garbage collector operates in two phases: In the mark phase, starting from a known root set of objects, the GC follows pointers in the object graph, marking each encountered object as alive. In the sweep phase, all objects not being alive are destroyed and all the live flags are reset.

In a process with many objects, having to traverse the whole graph may take a considerable amount of time. To mitigate this, a generational garbage collector extends this algorithm by assuming that the longer an object is alive, the lower the chances are that it will become unused. So, the objects are partitioned into different generations. The youngest generation will always be traversed while the older generations will be handled much less frequently or even never more at all.

@@

In 2010, Nat Tuck began working on a \textit{green\_threads} branch during his Google Summer of Code internship. The feature got prototyped using pure PIR and then implemented in Parrot's core. He got it to work in simple cases and started to work on operating system thread support but the internship ended before the code was ready to be merged into the master branch. The code lay dormant until the work on this thesis started in 2011.

@@

Green threads are one way to model concurrent control flows in a program. To get a better understanding of what they are and how they work, the possible options are discussed first.

@@

Coroutines are functions retaining their state between calls. Instead of returning, they yield control back to the calling function, possibly with an intermediate result returned. A very simple example in Python looks like:
\begin{GenericCode}
def counter():
    for i in range(0, 100):
	yield i
\end{GenericCode}
This function would return one integer for every call, counting from 0 to 100. An example usage would look like:
\begin{GenericCode}
x = counter()
print x
x = counter()
print x
\end{GenericCode}
This example would print the numbers 0 and 1. For the two functions to run in parallel, they have to cooperate. One by calling the other repeatedly and the callee by yielding control back to the caller.

@@

OS threads are like multiple processes running at the same time but share their memory. The operating system is responsible for giving each of these threads CPU time to run. OS threads are the only of these options where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.

@@

"Green threads" or "lightweight threads" are threads managed by the VM instead of by the operating system. For better distinction green threads are henceforth referred to as "tasks". These tasks only run pseudo parallel. The VM contains a scheduler and support for preempting running tasks.
Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
This is an implementation of the many-to-one threading model very similar to what an operating system running on a system with a single CPU core does.
%
Advantages of green threads are:
%
\begin{itemize}
\item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
\item Green threads are very light weight with low memory overhead and close to zero creation time.
\item They do not depend on OS threading support
\item The interpreter controls the point of preemption of a green thread.
\item The interpreter controls the exact scheduling policy and may allow the user to influence it or even take over completely.
\item Garbage collection can be implemented like in a single threaded process.
\item Critical sections can be protected by disabling the scheduler until the code is clear of the section.
\end{itemize}
%
Disadvantages include:
%
\begin{itemize}
\item They do not allow more than one CPU core to be used for computations.
\item Blocking calls like I/O block the whole interpreter including other green threads.
\item The interpreter has to contain logic and timers to control green threads.
\end{itemize}

Green threads differ from coroutines in that the interpreter decides when a running task is to be preempted while a coroutine depends on manual yield calls.

Following is a simple example to explain why concurrent processing can endanger the interpreter's internal consistency. The basic problem is concurrent writes to shared data. Assuming an implementation of an array class consisting of the field holding the data and the number of contained elements in a separate member variable:
\begin{CCode}
pmclass ResizableIntegerArray auto_attrs provides array {
    ATTR INTVAL   size;      /* number of INTVALs stored in this array */
    ATTR INTVAL * int_array; /* INTVALs are stored here */
\end{CCode}
to append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
Now if two threads simultaneously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size. In this case one of the appended values would be lost.

In a more complicated example, the array could have to resize its data buffer to accommodate the new value. It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one. Again if two threads try to do this simultaneously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer. This would lead to the copying thread accessing freed memory.

When garbage collection is brought into the mix, the possibilities for corruption grow even more. At the point where the first thread needs to allocate the new buffer, the garbage collector could decide that it needs to run and clear some unused memory. It would have to traverse the object tree to mark alive objects. Between the mark and the sweep runs, the second thread could change the pointer to the buffer from the old one to the new one. So the old buffer would have been marked alive, while during the sweep run the new buffer would be in its place, but not yet marked alive.

@@

Parrot's green threads implementation is based on Nat Tuck's green\_threads branch developed during his Google Summer of Code internship.

In Parrot green threads are called \textit{Task}. Each task is assigned a fixed amount of execution time. On expiry a timer callback sets a flag which is checked for at execution of every branch operation. Since the interpreter's state is well defined at this point, its internal consistency is guaranteed. The same goes for the garbage collector. Since task preemption is only done while executing user level code, the GC can do its work undisturbed and without the need for measures like locking. Since user level code is allowed to disable the scheduler, it can be guaranteed to run undisturbed through critical sections.

The scheduler is implemented as a PMC (Parrot's polymorphic container) type. The plan is to allow the user to subclass this PMC thus allowing very fine grained control over the scheduling policy. Features a user could add this way would be for example giving different priorities to tasks or implementing the possibility to suspend and resume a task.

@@

This chapter describes how green threads were used to solve the following problems occurring when trying to implement threading support:
%
\begin{itemize}
\item How to ensure internal interpreter consistency when doing writes to shared data.
\item How to implement critical sections.
\item How to handle garbage collection.
\end{itemize}

@@

As described in the introduction, cross thread writes to shared data may endanger the internal consistency of the interpreter. Traditionally the response to this problem is the use of locks of varying granularity. Fine grained locking allows code to run well parallelized but taking and releasing locks also costs performance. It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate. Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment \cite{LockingInOSKernels}. While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the very common case of having only a single thread running.

Too coarse locking on the other hand would reduce scalability and the performance gains through parallel execution by having threads wait for extended periods for locks to become open. In the extreme case of having a global interpreter lock it would even effectively serialize all computations costing much of the benefits of using threads in the first place.

The other problem with locking is the possibility of introducing dead locks. For example, two functions F1 and F2 both use two resources A and B protected by locks. If F1 first locks A and then tries to lock B while F2 has already locked B and is now trying to lock A, the program would come to a halt. Both functions would be left waiting for the other to unlock the resource which will never happen. With fine grained locking the possibilities for such bugs grow quickly. At the same time it is easy to miss a case where a lock would be appropriate leading to difficult to diagnose corruption bugs.

The solution for these problems implemented for this thesis is to side step them altogether by disallowing write access to shared data. The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started. The interpreter would then create proxy objects for these variables which the task can use to access the data. These proxies contain references to the original objects. They use these references to forward all reading \textit{vtable} functions to the originals. Write access on the other hand would lead to a runtime error.

In other words all data is owned by the thread creating it and only the owner may write to it. Other threads have only read access.

For threads to be able to communicate with their creators and other threads, they still need to write to shared data. This is where green threads come into play. Since green threads are very light weight, it is feasible for a thread to create a task just for updating a variable. This task would be scheduled on the interpreter owning this variable. To reduce latency, the task is flagged to run immediately. The data owning interpreter would preempt the currently running task and process the new write task. Put another way, the data owning interpreter is told what to write to its variables, so other threads don't have to.

@@

In this scheme critical sections can be implemented by disabling preemption till the code is clear of the section. With other threads not being allowed to write to the data and the current task running uninterrupted it it guaranteed to complete without the data being changed from outside.

\section{Garbage Collection}

A garbage collector has to traverse and process the entire tree of objects in memory. If during this traversal another thread is changing this tree by creating new objects, moving them around in the tree or removing them, the garbage collector's data would be compromised. For example it could happen that the GC does not mark referenced objects alive because they were added to parents having already been processed.

In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active. While multithreaded garbage collection algorithms exist, their implementation would have been too complex and brittle for this thesis \cite{VCGC}. Other VMs solve this problem by suspending all threads while the GC is running. Suspending all threads is not as simple as it sounds because these threads can be in any state at the time. For example, they could be blocking on some long running I/O operation like waiting for a reply on a network connection. A thread can stay in this condition indefinitely thereby never being able to confirm synchronization.

Since cross threaded writes are already forbidden and all read access to other thread's data goes through the very narrow channel of proxy objects, forcing a complete separation of the thread's memory domains was only a small step. By having separate memory areas for each interpreter, it became possible to have each interpreter to run its own garbage collector. This way, the known to work, single threaded GC implementation can be used nearly unchanged.

With separated memory domains it can happen that an object is created on one thread, used on another thread but not any more on the owner thread. Without any additional measures, the GC would collect such objects, since it does not know about references from other threads. But since all objects which could be accessed from other threads have to be pushed onto the Task object representing these threads, the objects can still be referenced from the Task object. Since the task is still owned by the original thread, the GC knows that these objects are still in use.

@@

\section{The Scheduler}

The scheduler is the place where most of the green thread logic is implemented. It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the Scheduler PMC type containing the scheduler's data and public interface.

\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g. on interpreter startup and whenever a new thread is created. It creates the scheduler PMC and sets up alarm signal handling.

The scheduler is hooked into the system by replacing the call to execute the \textit{main\_sub} by a call to \textit{Parrot\_cx\_begin\_execution}. The latter creates the main task using \textit{main\_sub} as the task's code. This task is then put onto the run queue. The scheduler timer gets enabled and control given to \textit{Parrot\_cx\_outer\_runloop}.

\textit{Parrot\_cx\_outer\_runloop} is the core loop taking tasks from the run queue and executing them. Despite its name, it should not be confused with runloops discussed in section~\ref{sec:Runloops}. \textit{Parrot\_cx\_outer\_runloop} actually consists of two nested loops. The inner loop fetches tasks and executes as long as tasks are available. Even when all tasks are finished there may still be alarms pending which upon expiry would trigger new tasks to be scheduled. For this reason there's an outer loop checking for pending alarms and if there are putting the thread to sleep until the next alarm expires.

\textit{Parrot\_cx\_next\_task} contains the code to take the next task from the run queue and executing it. Before calling the task, it checks if there are other tasks still in the queue. Only if other tasks are waiting, task preemption gets enabled. Otherwise the current task would run until it finishes by itself, schedules other tasks or an alarm expires. This is an optimization reducing the runtime overhead of green threads in the important single tasking case to 0.

To enable task preemption, \textit{Parrot\_cx\_enable\_preemption} sets a flag on the scheduler PMC and uses \textit{Parrot\_cx\_set\_scheduler\_alarm} to set an alarm at a point \textit{PARROT\_TASK\_SWITCH\_QUANTUM} milliseconds in the future.

When an alarm is triggered, Parrot only increments the global \textit{alarm\_serial}. This serial is checked for changes by \textit{Parrot\_alarm\_check} being called by \textit{Parrot\_cx\_check\_scheduler}. This function in turn is called once for every \textit{branch} op. When the serial is different from the last check or the \textit{SCHEDULER\_wake\_requested} flag is set, \textit{Parrot\_cx\_check\_scheduler} wakes the scheduler by calling \textit{Parrot\_cx\_run\_scheduler}.

Since POSIX alarms can only be delivered to the main thread of a process, \textit{Parrot\_cx\_run\_scheduler} uses \textit{Parrot\_thread\_notify\_threads} to relay alarms to other thread's schedulers. For this purpose every thread has a pipe which \textit{Parrot\_thread\_notify\_threads} writes a dummy byte to. The next step for \textit{Parrot\_cx\_run\_scheduler} is to check for any expired alarms using \textit{Parrot\_cx\_check\_alarms}. An alarm contains a task as callback. On expiry this callback gets scheduled for immediate execution. As alarm signals do not contain any information about their origin, the scheduler does not know if the alarm was scheduled by user code or if it is the scheduler's preemption alarm. \textit{Parrot\_cx\_check\_quantum} is therefore used to check if the current task has used its assigned quantum and if it should be preempted.

Preemption of a task is implemented in \textit{Parrot\_cx\_preempt\_task}. It uses \textit{Parrot\_cx\_stop\_task} to create a continuation at the current point of execution and store it as the task's code. The task itself is then appended at the end of the run queue. The NULL opcode return value is propagated all the way up through the stack back to the \textit{branch} op checking for alarms. The op then recognizes this as the signal to stop processing and to exit the runloop.

@@

\textit{Parrot\_cx\_schedule\_immediate} is used in various places like at alarm expiry for putting a task to the head of the run queue and immediately causing preemption of the currently running task. It does this by setting \textit{SCHEDULER\_wake\_requested} and \textit{SCHEDULER\_resched\_requested} flags the same as when the preemption alarm expires. To be precise this mechanism leads to the current task being preempted at the next \textit{branch} op since this is the place when the mentioned flags will be checked.

Alarms can be registered using \textit{Parrot\_cx\_schedule\_alarm}. It puts the alarm in the appropriate place in the ordered alarm list and uses \textit{Parrot\_alarm\_set} which sets the actual alarm but only if there's not another alarm already set for an earlier time since there can only be one POSIX alarm pending at any time.

\textit{Parrot\_cx\_schedule\_sleep} is the actual implementation of the \textit{sleep} op. Like in \textit{Parrot\_cx\_preempt\_task}, \textit{Parrot\_cx\_stop\_task} is used to get an updated task at the current execution position, but instead of pushing this on the run queue, it is used as a callback for a newly set alarm. Again the NULL opcode is used to stop processing of the current task.

@@

The Scheduler PMC contains the following attributes:
%
\begin{itemize}
\item PMC *task\_queue: List of tasks/green threads waiting to run
\item Parrot\_mutex task\_queue\_lock: a lock protecting the task\_queue so other threads can access it safely
\item PMC *alarms: List of future alarms ordered by time
\item PMC *all\_tasks: Hash of all active tasks by ID
\item UINTVAL next\_task\_id: ID to assign to the next created task
\item Parrot\_Interp interp: A reference to the scheduler's interpreter.
\end{itemize}

\textit{push\_pmc} and \textit{unshift\_pmc} are used to add a task at the end respectively the beginning of the task\_queue. \textit{shift\_pmc} is used to fetch and remove a task from the tip of the \textit{task\_queue}. \textit{get\_integer} returns the number of tasks in the \textit{task\_queue}. These four methods use the \textit{task\_queue\_lock} to make accessing the \textit{task\_queue} thread safe. Thus the Scheduler PMC acts as a container while using \textit{task\_queue} to actually store the data.

\textit{alarms} is list containing all pending alarms, sorted by their expiry time. Sorted insertion is used to keep alarms in order. Sorting makes it simple and efficient to retrieve all alarms that have expired at a certain point in time.

The \textit{active\_tasks} returns an array containing all tasks which have been run at least once and are not yet finished, e.g. they are currently being executed or are preempted.

@@

The Task PMC contains the following attributes:
%
\begin{itemize}
\item UINTVAL id: Unique identifier for the task
\item FLOATVAL birthtime: The creation time stamp of the task
\item Parrot\_Interp interp: The interpreter which created the task
\item PMC *code: The code to run
\item PMC *data: Additional data for the task given as parameter to \textit{code}
\item INTVAL killed: Flag marking killed tasks
\item PMC *waiters: Tasks waiting on this one
\item PMC *shared: List of variables shared with this task
\item PMC *partner: Copy of this task on the other side of a GC barrier, meaning in another thread
\end{itemize}

The \textit{invoke} method is used to run the task. It first checks the \textit{killed} flag to see if the task has been killed while waiting in the task queue. If it is still alive, the current recursion depth is saved to a local variable. This is a workaround needed because \textit{Parrot\_Sub\_invoke}, which ends up being called when invoking the task's \textit{code}, increments the recursion depth, but it does not get decremented anymore in case the task got preempted. This would lead to an ``recursion depth exceeded'' exception when many tasks are invoked during a program's runtime. The next steps are to add the task to the list of active tasks and to invoke the task's \textit{code}. \textit{data} is an optional parameter to the subroutine given as \textit{code}. Since this can be a compound object, a single parameter can cover all use cases.

@@

\textit{push\_pmc} and \textit{pop\_pmc} are used to add or remove variables to the task's list of shared data. For these variables, proxies will be created when the task is run on a different thread.

@@

Runloops are created whenever a part of Parrot's C internals starts to execute bytecode. This can be at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback. An exception handler can also be such a callback. Each runloop has a unique id. These ids are numbered from 0 and increased by 1 whenever a new runloop is created.

>>>>

This thesis shows a way to implement threading support in Parrot using a hybrid approach using a combination of lightweight (``green'') threads and OS threads. These lightweight threads are used as messages in a system where reading shared variables is allowed but only the one owner thread may write to it. The implementation of this hybrid approach is described in detail and benchmarks are presented demonstrating the viability of this design.

@@

In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared variables are expensive and still not all features of the language are usable in threaded programs.

@@

A mark and sweep GC operates in two phases: In the mark phase, starting from a known root set of objects, the GC follows pointers in the object graph, marking each encountered object as alive. In the sweep phase, all objects not being alive are destroyed and all the live flags are reset.

In a process with many objects, having to traverse the whole graph may take a considerable amount of time. To mitigate this, a generational GC extends this algorithm by assuming that the longer an object is alive, the lower the chances are that it will become unused. So, the objects are partitioned into different generations. The youngest generation will always be traversed while the older generations will be handled much less frequently or even never more at all.

@@

In 2010, Nat Tuck began working on a \textit{green\_threads} branch during his Google Summer of Code internship. The feature got prototyped using pure PIR and then implemented in Parrot's core. He got it to work in simple cases and started to work on OS thread support but the internship ended before the code was ready to be merged into the master branch. The code lay dormant until the work on this thesis started in 2011.

@@

Green threads are one way to model concurrent control flows in a program. To get a better understanding of what they are and how they work, the possible options for supporting concurrency are discussed before explaining green threads in detail.

@@

Coroutines are functions retaining their state between calls. Instead of returning, they yield control back to the calling function, possibly with an intermediate result returned. A very simple example in Python looks like:
\begin{GenericCode}
def counter():
    for i in range(0, 100):
        yield i
\end{GenericCode}
This function returns one integer for every call, counting from 0 to 100. An example usage looks like:
\begin{GenericCode}
def count\_up():
    x = counter()
    print x
    x = counter()
    print x
\end{GenericCode}
This example prints the numbers 0 and 1. For the two functions to run in parallel, they have to cooperate. One by calling the other repeatedly and the callee by yielding control back to the caller.

@@

OS threads are like multiple processes running in parallel but share their memory. The OS is responsible for giving each of these threads CPU time to run. OS threads are the only option where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.

@@

\textit{Green threads} or \textit{lightweight threads} are threads managed by the VM instead of by the OS. For better distinction green threads are henceforth referred to as \textit{tasks}. These tasks only run pseudo parallel. The VM has a scheduler and support for preempting running tasks.
Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
This is an implementation of the many-to-one threading model very similar to what an OS running on a system with a single CPU core does.
%
Advantages of green threads are:
%
\begin{itemize}
\item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
\item Green threads are light weight having low memory overhead and close to zero creation time.
\item They do not depend on OS threading support.
\item The interpreter controls the point of preemption of a green thread.
\item The interpreter controls the scheduling policy and may allow the user to influence it or even take over completely.
\item Garbage collection can be implemented like in a singlethreaded process.
\item Critical sections can be protected by disabling the scheduler until the code is clear of the section.
\end{itemize}
%
Disadvantages include:
%
\begin{itemize}
\item Green threads do not allow more than one CPU core to be used for computations.
\item Blocking calls like I/O block the interpreter including other green threads.
\item The interpreter has to provide logic and timers to control green threads.
\end{itemize}

Green threads differ from coroutines in that the interpreter decides when a running task is to be preempted while a coroutine depends on explicit yield calls.

Below is an example in order to explain why concurrent processing can endanger the interpreter's internal consistency. The basic problem are concurrent writes to shared variables. Assuming an implementation of an array class consisting of the field holding the data and the number of contained elements in a separate member variable:
\begin{CCode}
pmclass ResizableIntegerArray auto_attrs provides array {
    ATTR INTVAL   size;      /* number of INTVALs stored in this array */
    ATTR INTVAL * int_array; /* INTVALs are stored here */
\end{CCode}
To append a new value, the array has to read the current size, write the new value at the position \textit{size} + 1 and then write the incremented size back into the member variable.
Now, if two threads simultaneously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size. In this case one of the appended values is lost.

In a more complicated example, the array could have to resize its data buffer to accept the new value. It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroy the old one. Again, if two threads try to do this simultaneously, one of them could still be copying data, while the other already destroys the old buffer. This leads to the copying thread accessing freed memory.

When garbage collection is brought into the mix, the possibilities for corruption grow even further. At the point where the first thread needs to allocate the new buffer, the GC could decide that it needs to clear some unused memory. It would have to traverse the object graph to mark alive objects. Between the mark and the sweep phases, the second thread could change the pointer to the buffer from the old one to the new one. So the old buffer would have been marked alive, while during the sweep phase the new buffer would be in its place, but not yet marked alive.

@@

Parrot's green threads implementation is based on Nat Tuck's \textit{green\_threads} branch developed during his Google Summer of Code internship \cite{ChandonBlog}.

In Parrot, green threads are called \textit{Tasks}. Each task is assigned a fixed amount of execution time. After this time is up a timer callback sets a flag which is checked at execution of every branch operation. Since the interpreter's state is well defined at this point, its internal consistency is guaranteed. The same holds for the GC. Since task preemption is only done while executing user-level code, the GC can do its work undisturbed and without the need for measures like locking. Since user-level code is allowed to disable the scheduler, it can be guaranteed to run undisturbed through critical sections.

The scheduler is implemented as a PMC type. This allows the user to subclass this PMC thus allowing fine-grained control over the scheduling policy. Features, a user could add this way would be for example giving different priorities to tasks or implementing the possibility to suspend and resume a task.

@@

This chapter describes how green threads are used to solve the following problems occurring when trying to implement threading support:
%
\begin{itemize}
\item How to ensure internal interpreter consistency when doing writes to shared variables?
\item How to implement critical sections?
\item How to handle GC?
\end{itemize}

@@

As described in the introduction, cross-thread writes to shared variables may endanger the internal consistency of the interpreter. Traditionally, the solution to this problem is the use of locks of varying granularity. Fine-grained locking allows code to run in parallel but taking and releasing locks costs performance. It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate. Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment \cite{LockingInOSKernels}. While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the common case of having only a singlethread running.

Too coarse locking on the other hand would reduce scalability and the performance gains through parallel execution by having threads wait for extended periods for locks to become available. In the extreme case of having a global interpreter lock it would effectively serialize all computations costing much of the benefits of using threads in the first place.

The other problem with locking is the possibility of introducing deadlocks. For example, two functions \textit{F1} and \textit{F2} both use two resources \textit{A} and \textit{B} protected by locks. If \textit{F1} first locks \textit{A} and then tries to lock \textit{B} while \textit{F2} has already locked \textit{B} and is now trying to lock \textit{A}, the program would come to a halt. Both functions would be left waiting for the other to unlock the resource which will never happen. With fine-grained locking, the possibilities for such bugs grow quickly. At the same time, it is easy to miss a case where a lock would be appropriate leading to difficult to diagnose corruption bugs.

The solution for these problems implemented in this thesis is to side step them altogether by disallowing write access to shared variables. The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started. The interpreter would then create proxy objects for these variables which the task can use to access the data. These proxies contain references to the original objects. They use these references to forward all reading \textit{vtable} functions to the originals. Write access on the other hand would lead to a runtime error.

In other words, all data is owned by the thread creating it and only the owner may write to it. Other threads have only read access.

For threads to be able to communicate with their creators and other threads, they still need to write to shared variables. This is where green threads come into play. Since green threads are light weight, it is feasible for a thread to create a task just for updating a variable. This task is scheduled on the interpreter owning this variable. To reduce latency, the task is flagged to run immediately. The data-owning interpreter will preempt the currently running task and process the new write task. Put another way, the data-owning interpreter is told what to write to its variables, so other threads don't have to.

@@

Critical sections can be implemented by disabling preemption until the code has left the section. With other threads not being allowed to write to the data and the current task running uninterrupted it it guaranteed to complete without the data being changed from outside.

\section{Garbage Collection (GC)}

A GC has to traverse and process the entire graph of objects in memory. If during this traversal another thread is changing this structure by creating new objects, moving them around in the graph or removing them, the GC's data would be compromised. For example, it could happen that the GC does not mark referenced objects alive because they were added to objects which have already been processed.

In previous attempts to implement threading support in Parrot, the solution to this problem was to disable the GC while multiple threads were active. While multithreaded GC algorithms exist, their implementation would have been too complex and brittle for this thesis \cite{VCGC}. Other VMs solve this problem by suspending all threads while the GC is running. Suspending all threads is not as simple as it sounds because these threads can be in any state at the time. For example, they could be blocking on some long running I/O operation like waiting for a reply on a network connection. A thread can stay in this condition indefinitely thereby never being able to confirm synchronization.

Since cross-threaded writes are already forbidden and all read access to other thread's data goes through the narrow channel of proxy objects, forcing a complete separation of the thread's memory domains is only a small step. By having separate memory areas for each interpreter, it becomes possible to have each interpreter to run its own GC. This way, the known to work singlethreaded GC implementation can be used nearly unchanged.

With separated memory domains, it can happen that an object is created on one thread, used on another thread but not any more on the owner thread. Without any additional measures, the GC would collect such objects, since it does not know about references from other threads. But since all objects which could be accessed from other threads have to be pushed onto the task object representing these threads, the objects can still be referenced from the task object. Since the task is still owned by the original thread, the GC knows that these objects are still in use.

@@

\section{Scheduler}

The scheduler is the place where most of the green thread logic is implemented. It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the scheduler PMC type containing the scheduler's data and public interface.

\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g., on interpreter startup and whenever a new thread is created. It creates the scheduler PMC and sets up alarm signal handling.

The scheduler is hooked into the system by replacing the call to execute the \textit{main\_sub} by a call to \textit{Parrot\_cx\_begin\_execution}. The latter creates the main task using \textit{main\_sub} as the task's code. This task is then put onto the run queue. The scheduler timer gets enabled and control is given to \textit{Parrot\_cx\_outer\_runloop}.

\textit{Parrot\_cx\_outer\_runloop} is the loop taking tasks from the run queue and executing them. Despite its name, it may not be confused with runloops discussed in section~\ref{sec:Runloops}. \textit{Parrot\_cx\_outer\_runloop} actually consists of two nested loops. The inner loop fetches tasks from the task queue and executes as long as tasks are available. Even when all tasks are finished there may still be alarms pending which upon termination would trigger new tasks to be scheduled. For this reason, there is an outer loop checking for pending alarms. If it finds any it puts the thread to sleep until the next alarm expires.

\textit{Parrot\_cx\_next\_task} contains the code to take the next task from the run queue and executing it. Before calling the task, it checks if there are other tasks still in the queue. Only if other tasks are waiting, task preemption is enabled. Otherwise, the current task will run until it finishes by itself, schedules other tasks or an alarm expires. This is an optimization reducing the runtime overhead of green threads in the important single tasking case to zero.

To enable task preemption, \textit{Parrot\_cx\_enable\_preemption} sets a flag on the scheduler PMC and uses \textit{Parrot\_cx\_set\_scheduler\_alarm} to set an alarm at a point in time \textit{PARROT\_TASK\_SWITCH\_QUANTUM} milliseconds in the future.

When an alarm is triggered, Parrot only increments the global \textit{alarm\_serial}. This serial is checked for changes by \textit{Parrot\_alarm\_check} being called by \textit{Parrot\_cx\_check\_scheduler}. This function in turn is called once for every \textit{branch} operation. When the serial is different from the last check or the \textit{SCHEDULER\_wake\_requested} flag is set, \textit{Parrot\_cx\_check\_scheduler} wakes up the scheduler by calling \textit{Parrot\_cx\_run\_scheduler}.

Since POSIX alarms can only be delivered to the main thread of a process, \textit{Parrot\_cx\_run\_scheduler} uses \textit{Parrot\_thread\_notify\_threads} to relay alarms to other thread's schedulers. For this purpose, every thread has a pipe which \textit{Parrot\_thread\_notify\_threads} writes a dummy byte to. The next step for the \textit{Parrot\_cx\_run\_scheduler} is to check for any expired alarms using \textit{Parrot\_cx\_check\_alarms}. An alarm has a task as callback. On expiry, this callback gets scheduled for immediate execution. As alarm signals do not contain any information about their origin, the scheduler does not know if the alarm was scheduled by user code or if it is the scheduler's preemption alarm. \textit{Parrot\_cx\_check\_quantum} is therefore used to check if the current task has used its assigned quantum and if it should be preempted.

Preemption of a task is implemented in \textit{Parrot\_cx\_preempt\_task}. It uses \textit{Parrot\_cx\_stop\_task} to create a continuation at the current point of execution and store it as the task's code. The task itself is then appended at the end of the run queue. The function returns a \textit{NULL} opcode which is propagated all the way up through the stack back to the \textit{branch} operation checking for alarms. The operation then recognizes this as the signal to stop processing and to exit the runloop.

@@

\textit{Parrot\_cx\_schedule\_immediate} is used in various places like at alarm expiry for putting a task at the head of the run queue and immediately causing preemption of the currently running task. It does this by setting \textit{SCHEDULER\_wake\_requested} and \textit{SCHEDULER\_resched\_requested} flags the same as when the preemption alarm expires. To be precise, this mechanism leads to the current task being preempted at the next \textit{branch} operation since this is the place where the mentioned flags will be checked.

Alarms can be registered using \textit{Parrot\_cx\_schedule\_alarm}. It puts the alarm in the appropriate place in the ordered alarm list and uses \textit{Parrot\_alarm\_set} which sets the actual alarm but only if there is not another alarm already set for an earlier time, since there can only be one POSIX alarm pending at any time.

\textit{Parrot\_cx\_schedule\_sleep} is the actual implementation of the \textit{sleep} op. Like in \textit{Parrot\_cx\_preempt\_task}, \textit{Parrot\_cx\_stop\_task} is used to get an updated task at the current execution position, but instead of pushing this on the run queue, it is used as a callback for a newly set alarm. Again, the \textit{NULL} opcode is used to stop processing of the current task.

@@

The Scheduler PMC contains the following attributes:
%
\begin{itemize}
\item \textit{PMC *task\_queue}: list of tasks/green threads waiting to run.
\item \textit{Parrot\_mutex task\_queue\_lock}: a lock protecting the task\_queue so other threads can access it safely.
\item \textit{PMC *alarms}: list of future alarms ordered by time.
\item \textit{PMC *all\_tasks}: hash of all active tasks by ID.
\item \textit{UINTVAL next\_task\_id}: ID to assign to the next created task.
\item \textit{Parrot\_Interp interp}: a reference to the scheduler's interpreter.
\end{itemize}

\textit{push\_pmc} and \textit{unshift\_pmc} are used to add a task at the end, respectively the beginning of the task\_queue. \textit{shift\_pmc} is used to fetch and remove a task from the tip of the \textit{task\_queue}. \textit{get\_integer} returns the number of tasks in the \textit{task\_queue}. These four methods use the \textit{task\_queue\_lock} to make accessing the \textit{task\_queue} thread safe. Thus, the Scheduler PMC acts as a container while using \textit{task\_queue} to actually store the data.

\textit{alarms} is list containing all pending alarms, sorted by their expiration time. Sorted insertion is used to keep alarms in order. Sorting makes it simple and efficient to retrieve all alarms that have expired at a certain point in time.

The \textit{active\_tasks} returns an array containing all tasks which have been run at least once and are not yet finished, e.g., they are currently being executed or are preempted.

@@

The Task PMC contains the following attributes:
%
\begin{itemize}
\item \textit{UINTVAL id}: Unique identifier for the task.
\item \textit{FLOATVAL birthtime}: the creation time stamp of the task.
\item \textit{Parrot\_Interp interp}: the interpreter which created the task.
\item \textit{PMC *code}: the code to run.
\item \textit{PMC *data}: additional data for the task given as parameter to \textit{code}.
\item \textit{INTVAL killed}: flag marking killed tasks.
\item \textit{PMC *waiters}: tasks waiting on this one.
\item \textit{PMC *shared}: list of variables shared with this task.
\item \textit{PMC *partner}: copy of this task on the other side of a GC barrier, meaning in another thread.
\end{itemize}

The \textit{invoke} method is used to run the task. It first checks the \textit{killed} flag to see if the task has been killed while waiting in the task queue. Parrot manages a \textit{recursion depth} counter which records, how many levels the call stack has. This counter is incremented by \textit{Parrot\_Sub\_invoke} for each call to a subroutine and decremented when a subroutine returns. While \textit{Parrot\_Sub\_invoke} is used to start or resume a task, the executed subroutine does not return when the task is preempted. This leads to \textit{recursion depth} growing until a \textit{recursion depth exceeded} exception is thrown. Therefore, as a workaround, the current \textit{recursion depth} is saved to a local variable and restored when the task stops executing.

The next steps are to add the task to the list of active tasks and to invoke the task's \textit{code}. \textit{data} is an optional parameter to the subroutine given as \textit{code}. Since this can be a compound object, a single parameter can cover all use cases.

@@

\textit{push\_pmc} and \textit{pop\_pmc} are used to add or remove variables to the task's list of shared variables. For these variables, proxies will be created when the task is run on a different thread.

@@

Runloops are created whenever a part of Parrot's C internals starts to execute bytecode. This can be at interpreter startup to start the actual program execution or when some library function needs to execute a user callback. An exception handler can also be such a callback. Each runloop has a unique id. These ids are numbered from 0 and increased by 1 whenever a new runloop is created.

When an exception is thrown, Parrot searches for a previously set up exception handler and executes it. As part of exception handling finalization, it cleans up, freeing any information no longer needed. The exception handler does not have to have been set up in the same runloop from which the exception has been thrown. Parrot uses the runloop id to identify such situations and free the nested runloops up to the level of the exception handler.

A continuation also contains the id of the runloop in which it has been created. When resuming a preempted task, this id is not the same as the one of the runloop created for executing its code. This leads to Parrot not finding the correct runloop when finalizing an exception.
